Supplement: Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-05-16 nips_2018.tex
cut a bunch of stuff - premature optimization

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by training a neural recognition model, $q(\cdot |\cdot )$,
during the compilation phase: $q$ is trained to approximate the
posterior over programs, $q(p|x)\propto\probability[p|x,\mathcal{D},\theta}]=\probability[x|p]\probability[p|\mathcal{D},\theta}]$,
  thus amortizing the cost of finding programs with high posterior probability.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
We represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles the Lisp programming language.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is
used in functional programming languages like OCaml.
Type variables are always written using lowercase Greek letters
and we write $\alpha\to \beta$ to mean a function that takes an input of type $\alpha$
and returns something of type $\beta$.
We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
has the type $\tau$.
For example, to describe the type of the identity function
we would say \code{(lambda (x) x)}$:\alpha\to \alpha$.
We say a type $\alpha$ \emph{unifies} with $\tau$ if every expression
$p:\alpha$ also satisfies $p:\tau$. Furthermore, the act of \emph{unifying}
a type $\alpha$ with $\tau$ is to introduce constraints on the type
variables of $\alpha$ to ensure that $\alpha$ unifies with $\tau$.
See Supplement for more detail on program representation.
With this notation in hand we now define DSLs:

\noindent\textbf{Definition: $(\mathcal{D},\theta)$.}
A DSL $\mathcal{D}$ is a set of typed $\lambda$-calculus expressions.
%\\\noindent\textbf{Definition.}
A weight vector $\theta$ for a DSL $\mathcal{D}$ is a vector of $|\mathcal{D}| + 1$ real numbers:
one number for each DSL primitive $e\in \mathcal{D}$, written $\theta_e$,
and a weight controlling the probability of a variable occurring in a program, $\theta_{\text{var}}$.

\begin{wrapfigure}{L}{0.5\textwidth}
  \begin{minipage}{0.49\textwidth}    
\begin{algorithm}[H]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
%%      \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
%%   \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
%%   \STATE \textbf{Output:} a program whose type unifies with $\tau$
%%   \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
%% \STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $(\mathcal{D},\theta)$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} \code{(lambda (}var\code{) }body\code{)}
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | p: \tau' \in \mathcal{D}\cup\mathcal{E}$ \\
      \hspace*{6.9em}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$ \\
      \hspace*{8.8em}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Unify $\tau$ with yield$(\tau')$.
   \STATE $\left\{\alpha_k \right\}_{k = 1}^K\gets\text{args}(\tau')$ 
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} \code{(}$e\;a_1\; a_2\; \cdots\; a_K$\code{)}
 \STATE\textbf{where:}
 \STATE yield$(\tau) = \begin{cases}
   \text{yield}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   \tau   &\text{ otherwise.}
 \end{cases}$ 
 \STATE  args$(\tau) = \begin{cases}
   [\alpha] + \text{args}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   []   &\text{ otherwise.}
 \end{cases}$
\end{algorithmic}
\end{algorithm}
\vspace{-1.8cm}    \end{minipage}
  \end{wrapfigure}Alg.~\ref{programGenerativeModel} is a procedure for drawing
samples from the generative model $(\mathcal{D},\theta)$.  In practice, we
enumerate programs in order of their probability under  Alg.~\ref{programGenerativeModel} rather than sample them.

@@

        \STATE $\mathcal{D},\theta\gets $induceDSL$(\{\mathcal{F}^{\theta}_x\cup\mathcal{F}^{q}_x\}_{x\in X})$  \hspace{1.1cm} \footnotesize{\textbf{(Compress)}}
        %% \STATE Define $Q_x(z) \propto \begin{cases}
        %%   \probability[x|z]\probability[z|\mathcal{D}_i,\theta_i]&x\in \mathcal{F}_x\\
        %%   0&x\not \in \mathcal{F}_x
        %% \end{cases}
        \ENDFOR
        \STATE \textbf{return} $\mathcal{D},\theta,q$
      \end{algorithmic}
    \end{algorithm}
  \end{minipage}
\vspace{-0.7cm}\end{wrapfigure}To appropriately score each proposed $\mathcal{D}$ we must reestimate the weight vector $\theta$.
Although this  may seem 
very similar to estimating the parameters of a probabilistic context free grammar,
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
our DSLs are context-sensitive due to the presence of variables
in the programs and also due to the polymorphic typing system.
In the Supplement we derive a tractable MAP estimator for $\theta$.
Putting all these ingredients together, Alg.~\ref{mainAlgorithm} describes how we combine program search,
recognition model training, and DSL induction.
\begin{figure*}
\centering  \begin{tabular}{ll}
    \toprule
    Example programs in frontiers&Proposed subexpression\\\midrule
    \begin{tabular}{l}
      \code{(lambda (a b) (fold b (cons "," a)}\\
      \phantom{\code{(lambda }}\code{(lambda (x z) (cons x z))))}\\
      \code{(lambda (a b) (fold a b (lambda (x z) (cons x z))))}
    \end{tabular}&
    \begin{tabular}{l}
      \code{(fold a b (lambda (x z) }\\
      \phantom{\code{(fold a b }}\code{(cons x z)))}
      \end{tabular}
\\\bottomrule
  \end{tabular}
  \caption{The DSL induction algorithm proposes subexpressions of programs to add to the DSL.
    These subexpressions are taken from programs in the frontiers (left column), and can introduce new variables (right column: \code{a} and \code{b}). Here, the proposed subexpression appends two lists.} 
  \label{fragmentExample}
\end{figure*}

>>>>

Searching for programs is hard because
of the large combinatorial search space. We ease this difficulty by training a neural recognition model, $q(\cdot |\cdot )$,
during the compilation phase: $q$ is trained to approximate the
posterior over programs, $q(p|x)\approx \probability[p|x,\mathcal{D},\theta}]\propto\probability[x|p]\probability[p|\mathcal{D},\theta}]$,
  thus amortizing the cost of finding programs with high posterior probability.

@@

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
We represent programs as $\lambda$-calculus expressions.
$\lambda$-calculus is a formalism for expressing functional programs
that closely resembles the Lisp programming language.
$\lambda$-calculus includes variables, function application, and the ability to create new functions.
Throughout this paper we will write $\lambda$-calculus expressions in Lisp syntax.
Our programs are all strongly typed.
We use the Hindley-Milner polymorphic typing system~\cite{pierce} which is
used in functional programming languages like OCaml.
%% Type variables are always written using lowercase Greek letters
%% and we write $\alpha\to \beta$ to mean a function that takes an input of type $\alpha$
%% and returns something of type $\beta$.
%% We use the notation $p:\tau$ to mean that the $\lambda$-calculus expression $p$
%% has the type $\tau$.
%% For example, to describe the type of the identity function
%% we would say \code{(lambda (x) x)}$:\alpha\to \alpha$.
%% We say a type $\alpha$ \emph{unifies} with $\tau$ if every expression
%% $p:\alpha$ also satisfies $p:\tau$. Furthermore, the act of \emph{unifying}
%% a type $\alpha$ with $\tau$ is to introduce constraints on the type
%% variables of $\alpha$ to ensure that $\alpha$ unifies with $\tau$.
%% See Supplement for more detail on program representation.
We now define DSLs:

\noindent\textbf{Definition: $(\mathcal{D},\theta)$.}
A DSL $\mathcal{D}$ is a set of typed $\lambda$-calculus expressions.
%\\\noindent\textbf{Definition.}
A weight vector $\theta$ for a DSL $\mathcal{D}$ is a vector of $|\mathcal{D}| + 1$ real numbers:
one number for each DSL primitive $e\in \mathcal{D}$, written $\theta_e$ and controlling the probability of primitive $e$ occurring in a program,
and a weight controlling the probability of a variable occurring in a program, $\theta_{\text{var}}$.

Together with its weight vector,
a DSL defines a distribution over programs, $\probability[p|\mathcal{D},\theta]$.
In the supplement, we define this distribution  
by specifying a procedure for drawing samples from $\probability[p|\mathcal{D},\theta]$. Care must be taken to ensure that programs are well-typed
and that variable scoping rules are obeyed.
With this distribution in hand, we search for programs by enumerating
$\lambda$-calculus expressions in decreasing order of their probability under
$(\mathcal{D},\theta)$.

@@

        \STATE $\mathcal{D},\theta\gets $induceDSL$(\{\mathcal{F}^{\theta}_x\cup\mathcal{F}^{q}_x\}_{x\in X})$  \hspace{1.1cm} \footnotesize{\textbf{(Compress)}}
        %% \STATE Define $Q_x(z) \propto \begin{cases}
        %%   \probability[x|z]\probability[z|\mathcal{D}_i,\theta_i]&x\in \mathcal{F}_x\\
        %%   0&x\not \in \mathcal{F}_x
        %% \end{cases}
        \ENDFOR
        \STATE \textbf{return} $\mathcal{D},\theta,q$
      \end{algorithmic}
    \end{algorithm}
  \end{minipage}
\vspace{-0.7cm}\end{wrapfigure}%% To appropriately score each proposed $\mathcal{D}$ we must reestimate the weight vector $\theta$.
%% Although this  may seem 
%% very similar to estimating the parameters of a probabilistic context free grammar,
%% for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
%% our DSLs are context-sensitive due to the presence of variables
%% in the programs and also due to the polymorphic typing system.
%% In the Supplement we derive a tractable MAP estimator for $\theta$.
Putting all these ingredients together, Alg.~\ref{mainAlgorithm} describes how we combine program search,
recognition model training, and DSL induction.
\begin{figure*}
\centering  \begin{tabular}{ll}
    \toprule
    Example programs in frontiers&Proposed subexpression\\\midrule
    \begin{tabular}{l}
      \code{(lambda (a b) (fold b (cons "," a)}\\
      \phantom{\code{(lambda }}\code{(lambda (x z) (cons x z))))}\\
      \code{(lambda (a b) (fold a b (lambda (x z) (cons x z))))}
    \end{tabular}&
    \begin{tabular}{l}
      \code{(fold a b (lambda (x z) }\\
      \phantom{\code{(fold a b }}\code{(cons x z)))}
      \end{tabular}
\\\bottomrule
  \end{tabular}
  \caption{The DSL induction algorithm proposes subexpressions of programs to add to the DSL.
    These subexpressions are taken from programs in the frontiers (left column), and can introduce new variables (right column: \code{a} and \code{b}). Here, the proposed subexpression appends two lists.} 
  \label{fragmentExample}
\end{figure*}

### 2018-05-16 supplement.tex
cut a bunch of stuff - premature optimization

>>>>

\section{Generative model over the programs}
\begin{wrapfigure}{L}{0.5\textwidth}
  \begin{minipage}{0.49\textwidth}    
\begin{algorithm}[H]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
%%      \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
%%   \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
%%   \STATE \textbf{Output:} a program whose type unifies with $\tau$
%%   \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
%% \STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $(\mathcal{D},\theta)$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} \code{(lambda (}var\code{) }body\code{)}
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | p: \tau' \in \mathcal{D}\cup\mathcal{E}$ \\
      \hspace*{6.9em}$\text{if }\tau\text{ can unify with yield}(\tau') \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$ \\
      \hspace*{8.8em}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Unify $\tau$ with yield$(\tau')$.
   \STATE $\left\{\alpha_k \right\}_{k = 1}^K\gets\text{args}(\tau')$ 
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} \code{(}$e\;a_1\; a_2\; \cdots\; a_K$\code{)}
 \STATE\textbf{where:}
 \STATE yield$(\tau) = \begin{cases}
   \text{yield}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   \tau   &\text{ otherwise.}
 \end{cases}$ 
 \STATE  args$(\tau) = \begin{cases}
   [\alpha] + \text{args}(\beta)   &\text{ if }\tau = \alpha\to \beta\\
   []   &\text{ otherwise.}
 \end{cases}$
\end{algorithmic}
\end{algorithm}
\vspace{-1.8cm}    \end{minipage}
  \end{wrapfigure}Alg.~\ref{programGenerativeModel} is a procedure for drawing
samples from the generative model $(\mathcal{D},\theta)$.  In practice, we
enumerate programs in order of their probability under  Alg.~\ref{programGenerativeModel} rather than sample them.
